---
title: "In-class Practice with the `scorecard` data"
author: "Sabrina Nardin"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,      # show code
  message = FALSE,  # suppress messages
  warning = FALSE   # (optional) suppress warnings too
)
```

This file includes solutions to today's in-class exercise, with explanations of code and graphs.
 
Load libraries and data:
```{r}
library(tidyverse)
library(rcis)
data(scorecard)
```

Explore data:
```{r}
glimpse(scorecard)
```


## Graph 1: Plot the total number of schools in the U.S. by school type

To represent one categorical variable, here `type`, the best graph is a **bar chart**. A bar chart takes each category of the categorical variable (here `"Public"`, `"Private nonprofit"`, `"Private for profit"`), and counts how many observations fall into each category. 
 
```{r}
ggplot(scorecard, mapping = aes(x = type)) +
  geom_bar() 
```

Interpretation question: What does this bar chart tell us about the number of schools in each school type?

**Bar chart "`stat = count`"**

```{r}
ggplot(data = scorecard, mapping = aes(x = type)) +
  geom_bar(stat = "count")
```

Note that `geom_bar()` automatically counts how many observations (rows) fall into each category and displays those counts on the y-axis. It does this because the default stat for `geom_bar()` is `stat = "count"`, not `stat = "identity"`.

Documentation question: Go online at https://ggplot2.tidyverse.org/reference/geom_bar.html or type `?geom_bar()` in the console to check the default `stat` for `geom_bar()`. What do you notice?

Code challenge:  Reproduce the same plot, but this time calculate the counts yourself first. Then tell `geom_bar()` to use those counts directly by setting stat = "identity". Hint: You must first calculate the counts for each school type. You can use the `count()` function for this, and then manually supply both `x` and `y` to your plot.

```{r}
# code challenge solution 

# first, calculate how many schools are in each type
scorecard_type_counts <- count(scorecard, type)

# look at the new summarized data
scorecard_type_counts

# then, make the bar chart using the summarized counts
ggplot(data = scorecard_type_counts,
       mapping = aes(x = type, y = n)) +
  geom_bar(stat = "identity")
```

Because we used `count()` first, the data are already summarized. The column `n` already contains the number of schools in each type. That is why we use `stat = "identity"`: we are telling ggplot to use the `n` column as it is, rather than counting the rows again.

Note: There is another geom for bar charts called `geom_col()`. Use it when your data already contains the numbers you want to show as bar heights (e.g., they are already counted).

```{r}
# Same plot using geom_col()
ggplot(data = scorecard_type_counts, 
       mapping = aes(x = type, y = n)) + 
  geom_col()
```

In short (for bar charts), use `geom_bar()` when you want ggplot to count rows for you, use `geom_col()` when the bar heights are already in your data. This is because technically, `geom_bar()` counts rows by default, while `geom_col()` uses the y-values you provide. This means that `geom_col()` is similar to writing `geom_bar(stat = "identity")`.


## Graph 2: Plot the annual total cost of school attendance across the U.S.

To represent one continuous variable, here `cost`, the best graph is a **histogram**. A histogram divides the input variable into "n" bins of equal width and counts how many observations fall into each bin. You can adjust the number of bins to see the distribution at different levels of detail.

```{r}
ggplot(data = scorecard, mapping = aes(x = cost)) +
  geom_histogram()
```

Interpretation questions: What does this histogram tell us about the distribution of annual college attendance costs? Where do most schools seem to fall? Are there any unusually high or low values?

Code questions: What are some alternative ways to rewrite this code and generate the same plot? Below I am proposing a few alternatives that are all valid. The original version and Alternative 1 are the most common. You can omit argument names such as `data =` and `mapping =`, but I recommend leaving them in to make the code easier to read.

```r
# Alternative 1
ggplot(data = scorecard) +
  geom_histogram(mapping = aes(x = cost))
```

```r
# Alternative 2
ggplot() +
  geom_histogram(data = scorecard, 
                 mapping = aes(x = cost))
```

```r
# Alternative 3 using `ggplot2` and `dplyr`
scorecard %>%
  ggplot(mapping = aes(x = cost)) +
  geom_histogram()
```

**Histograms "stat = `"bin"`**

One thing to notice is that the default stat for `geom_histogram()` is "bin", not "count" as with `geom_bar()`. This is because histograms work with continuous variables (whereas bar graphs work with categorical variables).

With a continuous variable, the categories do not already exist in the data like for categorical variables. Before ggplot can count anything, it first divides the x-variable into intervals or categories, called bins. Then it counts how many observations fall into each bin.

```{r}
ggplot(data = scorecard, mapping = aes(x = cost)) +
  geom_histogram(stat = "bin", 
                 bins = 30) # try changing the integer 
```

Each bar is a bin, which represents one interval of the data. For example, we can divide the data into 50, 30 (default), or 10 bins of equally sized bars. For each bin, `geom_histogram()` counts how many observations fall into that interval.

Explore the documentation: Go online or type `?geom_histogram()` in the console to check the default `stat` for `geom_histogram()`. What do you notice, and what does that mean?


**Adding title, labels, etc.**

```{r}
ggplot(data = scorecard, mapping = aes(x = cost)) +
  geom_histogram() +
  labs(
    title = "Distribution of Annual College Attendance Costs",
    x = "Published Annual Attendance Cost (USD)",
    y = "Number of Colleges",
    caption = paste("Data: College Scorecard (2018–2019; n =", nrow(scorecard), ")")
  ) +
  theme(
    plot.title = element_text(size = 16),
    plot.caption = element_text(size = 8, color = "gray")
  )
```

Tip: Titles, axis labels, and captions are important for finalized plots that you plan to present or share. During exploratory data analysis, it is often better to focus first on understanding the patterns: redraw the plot, aggregate the data, change the plot type, or adjust the variables. Once you know what you want to show, polish the final graph.


## Graph 3: Plot the annual total cost and net cost of attendance to schools in the U.S. 

To represent two continuous variables, here `cost` and `netcost`, the best graph is a **scatter plot**. Each point represents an observation. The position of a point on the x-axis corresponds to one variable, while the position on y-axis corresponds to the other variable.

```{r}
ggplot(scorecard,
  mapping = aes(x = cost, y = netcost)) +
  geom_point()
```

Interpretation questions: What does this scatter plot tell us about the relationship between advertised cost and net cost? What do you notice about the points that are not closely aligned? What additional exploratory analysis could we do next?

Example interpretation: As the advertised cost of attendance increases, the net cost also tends to increase, though there is substantial variation. Some schools have a much lower net cost than their advertised cost. The points do not fall neatly along a diagonal line, which suggests that net costs are often lower than advertised costs, especially for schools with higher advertised costs.


## Graph 4: Plot the annual total cost of attendance by school type 

To represent one continuous variable, here `cost`, and one categorical variable, here `type`, the best graph is a **box plot**. A box plot summarizes the distribution of the continuous variable across each category of the categorical variables.

```{r}
ggplot(scorecard, mapping = aes(x = type, y = cost)) +
  geom_boxplot()
```

Interpretation questions: What does this box plot tell us about cost differences across school types? How do we read the median, the box, the whiskers, and the outliers in this boxplot? See R for Data Science, 1st edition, Ch. 7: <https://d33wubrfki0l68.cloudfront.net/153b9af53b33918353fda9b691ded68cd7f62f51/5b616/images/eda-boxplot.png>.

Example interpretation: On average, public universities tend to have the lowest annual attendance costs, followed by private for-profit institutions. Private nonprofit institutions tend to have the highest costs. Useful follow-up questions you can further explore include: What might explain these differences? Are these patterns consistent with what you already know about U.S. higher education?

Other than using a boxplot, you can also use a **faceted histogram**, which creates a multiple panels, one for each school type. More formally, a faceted histogram looks at the distribution of a continuous variable (here `cost`) across different levels of the categorical variable (here `type`). The y-axis shows frequency count for each level of the categorical variable. Histograms cannot map a second variable directly, but we can use facets to compare the distribution of each college type.

```{r}
ggplot(scorecard,
  mapping = aes(x = cost)) +
  geom_histogram() +
  facet_wrap(facets = vars(type), nrow = 3) # try without nrow or with nrow = 1
```

Comparison question: Compare this faceted histogram with the boxplot. Which one do you prefer for this question, and why?


## Graph 5: Plot the total annual cost of attendance and net cost of attendance by school type

In this final example, we are plotting three variables at the same time: two continuous variables, here `cost` and `netcost`, and one categorical variable, here `type`. We can use a **faceted scatter plot**, which plots each category of the categorical variable (school type) in a separate panel with the same scale range on the x and y. 

```{r}
ggplot(scorecard,
  mapping = aes(x = cost, y = netcost)) +
  geom_point() +
  facet_wrap(facets = vars(type), nrow = 3) # try without nrow or with nrow = 1
```

Rather than using facets for school type, we can use the color aesthetic we learned in the previous lecture to show `type` in the same panel:
```{r}
ggplot(scorecard,
  mapping = aes(x = cost, y = netcost, color = type)) +
  geom_point()
```

Challenge: Try to add a fourth variable, `debt`, and represent it through the `size` aesthetic, for example `size = debt`. Does the plot become more informative, or harder to read? 
 
```{r}
# code challenge solution
ggplot(scorecard,
  mapping = aes(x = cost, 
                y = netcost, 
                color = type, 
                size = debt)) +
  geom_point()
```
 

This plot includes more information, but it is also harder to interpret. The color aesthetic already asks the reader to compare school types. Adding point size for debt adds another layer, but it can make the graph visually busy and harder to read. In general, it is often best not to include more than three variables in a single plot. Thus, a better approach is to create a separate plot focused only on debt, or to facet by school type and use size more selectively. 


## Other types of graphs

Today we examined several common graph types, but there are many other options to experiment with. The best graph depends on the type and number of variables you want to represent, and on the question you are trying to answer.

For more graph ideas, check today's readings on the course website and the visualization cheat sheet:

* Help > Cheat Sheets > Data Visualization with ggplot2
* Posit also has several cheat sheets for the tidyverse and R: <https://posit.co/resources/cheatsheets/>


